---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 96)
```

# ducksemantics

`ducksemantics` is a DuckDB-native semantic graph and grounding package for R.
The package owns the graph schema, OBO import, alias indexing, mention
grounding, provider interfaces, model judgment, and benchmark measurement.

HPO and MONDO are first-class graph sources here, but they are not the whole
abstraction. The same tables can hold ORPHANET, study notes, pi-bio-agent memory
edges, run ledgers, and local concept maps. BebeLM/Rbebelm is the bundled
on-device provider for CPU embedding and judgment; other providers implement
the same S7 generics.

```{r}
suppressPackageStartupMessages({
  library(ducksemantics)
  library(Rbebelm)
  library(Rfmalloc)
})

cache_dir <- tools::R_user_dir("ducksemantics", "cache")

ontology_files <- c(
  HPO = ducksemantics_cache_file(
    "https://purl.obolibrary.org/obo/hp.obo",
    "hp.obo",
    cache_dir = cache_dir
  ),
  MONDO = ducksemantics_cache_file(
    "https://purl.obolibrary.org/obo/mondo.obo",
    "mondo.obo",
    cache_dir = cache_dir
  )
)

weights_file <- Sys.getenv(
  "BEBELM_WEIGHTS_FILE",
  "/root/bebelm/LFM2.5-8B-A1B-Q4_K_M.gguf"
)
stopifnot(file.exists(weights_file))
```

The OBO parser produces the graph rows that DuckDB stores and indexes.

```{r}
graphs <- list(
  HPO = ducksemantics_read_obo(ontology_files[["HPO"]], family = "HPO", source = "hp.obo"),
  MONDO = ducksemantics_read_obo(ontology_files[["MONDO"]], family = "MONDO", source = "mondo.obo")
)

data.frame(
  source = names(graphs),
  nodes = vapply(graphs, function(x) nrow(x$nodes), integer(1)),
  aliases = vapply(graphs, function(x) nrow(x$aliases), integer(1)),
  edges = vapply(graphs, function(x) nrow(x$edges), integer(1)),
  row.names = NULL
)
```

The database below is rebuilt from those full ontologies, then indexed once for
lexical grounding.

```{r}
db_path <- file.path(cache_dir, "readme.duckdb")
unlink(db_path)

conn <- ducksemantics_connect(db_path)
ducksemantics_write_graph(
  conn,
  nodes = graphs$HPO$nodes,
  aliases = graphs$HPO$aliases,
  edges = graphs$HPO$edges,
  replace = TRUE,
  index = FALSE
)
ducksemantics_write_graph(
  conn,
  nodes = graphs$MONDO$nodes,
  aliases = graphs$MONDO$aliases,
  edges = graphs$MONDO$edges,
  index = TRUE
)

ducksemantics_index_stats(conn)$tables
```

Grounding returns deterministic candidates with source provenance. The diabetes
mention is deliberately negated; the lexical layer should surface it, and the
judgment layer should decide whether it belongs in the patient phenotype.

```{r}
note <- paste(
  "The proband presents with global developmental delay, seizures,",
  "short stature, cardiomyopathy, and no diabetes mellitus."
)

mentions <- ducksemantics_annotate(conn, note, document_id = "readme-case-001")
unique(mentions[, c("span", "node_id", "alias_kind", "source")])
```

The BebeLM bridge is typed as an embedding provider. The model load controls the
Rust thread pool; the provider controls token batching for long texts and
sequence batching across independent texts. The vectors below come from the
local GGUF model and are cached as RDS because they are deterministic for the
same model and pooling configuration.

```{r}
bebel_threads <- as.integer(Sys.getenv("BEBELM_NUM_THREADS", "16"))
model <- bebel_model_load(weights_file, num_threads = bebel_threads)
embedding_provider <- ducksemantics_bebel_embedding_provider(
  model,
  pooling = "mean",
  token_batch_size = 512L,
  sequence_batch_size = 1024L,
  check_interrupt = TRUE
)

embedding_terms <- c(
  "global developmental delay",
  "seizures",
  "short stature",
  "cardiomyopathy",
  "diabetes mellitus"
)

embeddings <- ducksemantics_cache_rds(
  file.path(cache_dir, "readme-bebel-embeddings.rds"),
  function() ducksemantics_embed(embedding_provider, embedding_terms)
)

data.frame(
  terms = nrow(embeddings),
  dimensions = ncol(embeddings),
  first_row_l2 = round(sqrt(sum(embeddings[1, ] ^ 2)), 4)
)
```

BebeLM embeddings are written back to DuckDB, then searched with DuckDB vector
functions. HNSW can be added by the same index spec; the README materializes the
fixed-dimension table without building the extension-backed index so the example
stays focused on the package API.

```{r}
embedding_label <- "Rbebelm mean"

ducksemantics_embedding_batch(
  embeddings = embeddings,
  subject_id = embedding_terms,
  subject_kind = "readme_term",
  provider = embedding_label,
  text = embedding_terms
) |>
  ducksemantics_write_embeddings(conn, replace = TRUE)

ducksemantics_embedding_query(
  embeddings[match("short stature", embedding_terms), , drop = FALSE],
  provider = embedding_label,
  subject_kind = "readme_term",
  top_k = 3
) |>
  ducksemantics_embedding_search(conn)

embedding_index <- ducksemantics_embedding_index_spec(
  dimensions = ncol(embeddings),
  provider = embedding_label,
  subject_kind = "readme_term",
  hnsw = FALSE
) |>
  ducksemantics_materialize_embedding_index(conn)

ducksemantics_embedding_query(
  embeddings[match("cardiomyopathy", embedding_terms), , drop = FALSE],
  table = embedding_index,
  provider = embedding_label,
  subject_kind = "readme_term",
  top_k = 3
) |>
  ducksemantics_embedding_search(conn)
```

The first semantic measurement is ontology clustering: embed every HPO and MONDO
label, store the vectors in DuckDB, cluster them, and compare cluster
assignments against direct graph edges. The embeddings are cached in chunks
because this is a real BebeLM pass over the ontology labels.

```{r}
ontology_label_terms <- DBI::dbGetQuery(
  conn,
  paste(
    "SELECT node_id, family, label",
    "FROM semantic_nodes",
    "WHERE label IS NOT NULL",
    "ORDER BY family, node_id"
  )
)

ontology_embedding_label <- "Rbebelm ontology label mean"
ontology_label_embeddings <- ontology_label_terms$label |>
  ducksemantics_embed_cached(
    provider = embedding_provider,
    cache_dir = file.path(cache_dir, "readme-bebel-ontology-label-embeddings"),
    chunk_size = 1024L
  )

ducksemantics_embedding_batch(
  embeddings = ontology_label_embeddings,
  subject_id = ontology_label_terms$node_id,
  subject_kind = "node",
  provider = ontology_embedding_label,
  text = ontology_label_terms$label
) |>
  ducksemantics_write_embeddings(conn, replace = TRUE)

rfmalloc_path <- file.path(cache_dir, "readme-ontology-clusters.fmalloc")
unlink(rfmalloc_path, force = TRUE)
rfmalloc_runtime <- Rfmalloc::open_fmalloc(
  rfmalloc_path,
  mode = "scratch",
  size_gb = 8
)

ontology_clusters <- ducksemantics_embedding_cluster_spec(
  k = 32,
  provider = ontology_embedding_label,
  subject_kind = "node",
  dimensions = ncol(ontology_label_embeddings),
  run_id = "readme-ontology-label-clusters",
  nstart = 1,
  max_iter = 30,
  storage = "rfmalloc"
) |>
  ducksemantics_cluster_embeddings(conn, rfmalloc_runtime = rfmalloc_runtime)

Rfmalloc::cleanup_fmalloc(rfmalloc_runtime)

head(ontology_clusters$summary[order(-ontology_clusters$summary$size), ], 8)
ducksemantics_embedding_cluster_graph_agreement(
  conn,
  ontology_clusters$cluster_run_id,
  predicates = c("is_a", "part_of")
)
```

BebeLM judgment uses the same provider interface. The parser here accepts
BebeLM tool calls and JSON judgment payloads, then returns the rows expected by
the semantic store.

```{r}
judgment_mentions <- data.frame(
  mention_id = "readme-judgment:000001",
  node_id = "HP:0004322",
  span = "short stature",
  start_offset = 26L,
  end_offset = 39L,
  score = 1,
  alias = "short stature",
  alias_kind = "label",
  source = "hp.obo",
  stringsAsFactors = FALSE
)

judgment_instructions <- c(
  "You adjudicate deterministic semantic grounding candidates.",
  "Return only JSON: an array of objects with mention_id, decision, confidence, evidence_span, short_reason.",
  "Use keep for a supported patient finding. Never invent identifiers."
)

judgment <- ducksemantics_cache_rds(
  file.path(cache_dir, "readme-bebel-judgment.rds"),
  function() {
    agent <- bebel_agent(model, greedy = TRUE, max_gen = 96, max_think = 0)
    ducksemantics_bebel_judge(
      agent,
      "The proband presents with short stature.",
      judgment_mentions,
      instructions = judgment_instructions,
      parser = ducksemantics_bebel_tool_judgment_parser(),
      record = FALSE
    )
  }
)

judgment[, c("subject_id", "object_id", "decision", "confidence", "model")]
```

Benchmarks are ordinary R data flowing through pipes. The metrics below use
gold HPO phenotypes and expose the current lexical baseline before model
judgment removes negated or disease-level mentions.

```{r}
suite <- ducksemantics_benchmark_cases(
  cases = data.frame(
    case_id = "readme-case-001",
    text = note,
    stringsAsFactors = FALSE
  ),
  gold = data.frame(
    case_id = "readme-case-001",
    node_id = c("HP:0001263", "HP:0001250", "HP:0004322", "HP:0001638"),
    stringsAsFactors = FALSE
  ),
  suite = "readme-hpo-mondo"
)

run <- suite |>
  ducksemantics_benchmark(conn)

run$metrics
run$timings[, c("case_id", "seconds", "token_count", "prediction_count", "prediction_bytes")]
```

```{r}
DBI::dbDisconnect(conn, shutdown = TRUE)
```
