---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

embeddinggemma_weights <- Sys.getenv("EMBEDDING_GEMMA_WEIGHTS_FILE", NA_character_)
colbert_weights <- Sys.getenv("COLBERT_WEIGHTS_FILE", NA_character_)
bebel_weights <- Sys.getenv("BEBELM_WEIGHTS_FILE", NA_character_)
stopifnot(
  file.exists(embeddinggemma_weights),
  file.exists(colbert_weights),
  file.exists(bebel_weights)
)
```

# ducksemantics

`ducksemantics` is a DuckDB-native semantic graph and evidence-grounding
package for R. It stores ontology nodes, aliases, graph edges, dense vectors,
and ColBERT document token vectors in one auditable DuckDB database.

It combines the Rbebelm retrieval and judgment APIs in one graph workflow.
EmbeddingGemma supplies task-prompted dense vectors for broad retrieval and
DuckDB HNSW candidate generation, LFM2.5-ColBERT supplies exact
late-interaction MaxSim reranking of stored candidate documents, and BebeLM
records structured judgments with negation, uncertainty, family history, and
subject context. HPO, MONDO, ORPHANET, and local knowledge graphs load through
the same graph schema.

## Install

Install the release and its R dependencies from the project R-universe:

```r
install.packages(
  "ducksemantics",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org")
)
```

## Build a semantic graph

The bundled HPO fixture is a small, attribution-preserving subset of the real
Human Phenotype Ontology. It keeps the README executable without downloading an
ontology during rendering.

```{r graph}
library(ducksemantics)

conn <- ducksemantics_connect()
hpo <- ducksemantics_read_obo(
  system.file("extdata", "hpo-readme.obo", package = "ducksemantics", mustWork = TRUE),
  family = "HPO",
  source = "HPO README fixture"
)
ducksemantics_write_graph(
  conn,
  nodes = hpo$nodes,
  aliases = hpo$aliases,
  edges = hpo$edges,
  replace = TRUE
)

# Deterministic, explainable candidates. This intentionally includes negated
# mentions; BebeLM judgment decides whether the mention is a patient finding.
case_text <- "The proband has short stature and seizures but no diabetes mellitus."
mentions <- ducksemantics_annotate(
  conn,
  case_text,
  document_id = "case-001"
)
mentions
```

### Full HPO validation

The 0.1.0 release was exercised against the complete HPO 2026-06-23 ontology.
The run loaded 19,836 active nodes, 50,029 aliases, and 24,378 direct edges,
then materialized 202,740 `is_a` closure rows. The model-scale run persisted one
256-dimensional EmbeddingGemma vector for every active term and 182,884 native
ColBERT token vectors across the same 19,836 terms before running dense and
exact MaxSim queries.

## Dense retrieval with EmbeddingGemma

EmbeddingGemma provides the broad candidate stage. Choose its task deliberately:
vectors sharing a compatible prompt contract can be compared directly. This
example loads the local GGUF selected by `EMBEDDING_GEMMA_WEIGHTS_FILE` and
executes the native encoder.

```{r dense}
library(Rbebelm)

embedding_model <- embeddinggemma_model_load(
  embeddinggemma_weights,
  num_threads = 2
)
dense_provider <- ducksemantics_embeddinggemma_provider(
  embedding_model,
  label = "embeddinggemma-hpo-v1",
  task = "semantic_similarity",
  dimensions = 256L
)

terms <- DBI::dbGetQuery(conn, "SELECT node_id, label FROM semantic_nodes WHERE family = 'HPO'")
vectors <- ducksemantics_embed_cached(
  terms$label,
  provider = dense_provider,
  cache_dir = file.path(tempdir(), "ducksemantics-embeddinggemma-hpo-cache")
)
ducksemantics_embedding_batch(
  vectors,
  subject_id = terms$node_id,
  subject_kind = "hpo_term",
  provider = "embeddinggemma-hpo-v1",
  text = terms$label
) |> ducksemantics_write_embeddings(conn, replace = TRUE)

# Exact DuckDB vector search is sufficient for this small example. An HNSW
# index is optional for a larger dense candidate collection.
dense_hits <- ducksemantics_embedding_query(
  ducksemantics_embed(dense_provider, "short height")[1L, ],
  provider = "embeddinggemma-hpo-v1",
  subject_kind = "hpo_term",
  top_k = 3L
) |> ducksemantics_embedding_search(conn)
dense_hits
```

## Exact late interaction with native ColBERT

ColBERT document vectors are persisted once, and each query is encoded with the
model's query contract. DuckDB then computes the exact MaxSim sum over stored
document vectors. Large corpora can send dense HNSW, lexical, or graph
candidates to this stage through `candidate_subject_id`. This example loads the
local GGUF selected by `COLBERT_WEIGHTS_FILE` and uses Rbebelm's native
LFM2.5-ColBERT encoder.

```{r colbert}
colbert_model <- colbert_model_load(colbert_weights, num_threads = 2)
colbert_documents <- ducksemantics_colbert_provider(
  colbert_model,
  role = "document",
  label = "lfm2.5-colbert-350m-q4km"
)

ducksemantics_token_embedding_batch_from_provider(
  terms$label,
  provider = colbert_documents,
  subject_id = terms$node_id,
  subject_kind = "hpo_term"
) |> ducksemantics_write_token_embeddings(conn, replace = TRUE)

hits <- ducksemantics_colbert_query(
  colbert_model,
  "developmental delay with seizures",
  provider = "lfm2.5-colbert-350m-q4km",
  subject_kind = "hpo_term",
  top_k = 25L
) |> ducksemantics_late_interaction_search(conn)
hits
```

The token table holds 128-dimensional LFM2.5-ColBERT document vectors for
variable-length MaxSim scoring. `duckdb-vss` can accelerate the fixed-width
EmbeddingGemma candidate stage while DuckDB evaluates the token-level reranker.

## BebeLM candidate judgment

BebeLM reviews candidates together with the retrieved context and records a
structured decision for negated, uncertain, historical, and subject-specific
mentions. Replacement concepts come from the supplied candidates and graph
neighbors, preserving a direct evidence trail. This example loads the local
GGUF selected by `BEBELM_WEIGHTS_FILE` and executes a BebeLM judgment.

```{r judge}
generation_model <- bebel_model_load(bebel_weights, num_threads = 2)
agent <- bebel_agent(generation_model, greedy = TRUE, max_gen = 256L, max_think = 0L)

judgments <- ducksemantics_bebel_judge(
  agent,
  text = case_text,
  mentions = mentions,
  conn = conn,
  parser = ducksemantics_bebel_tool_judgment_parser(),
  model = "LFM2.5-8B-A1B"
)
judgments
```

```{r cleanup, include=FALSE}
DBI::dbDisconnect(conn, shutdown = TRUE)
```
