---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
real_hpo_readme <- identical(tolower(Sys.getenv("RFASTHPOCR_README_REAL_HPO")), "true")
```

# RfastHPOCR

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/RfastHPOCR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/RfastHPOCR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`RfastHPOCR` is a reticulate-backed R binding to Tudor Groza's
[`FastHPOCR`](https://github.com/tudorgroza/fast_hpo_cr) Python package for
fast concept recognition over Human Phenotype Ontology terms in free text.

The package keeps the upstream Python implementation as the execution engine and
adds R-oriented helpers for:

- declaring and installing the Python dependency set (`FastHPOCR`, `pronto`, and
  `tqdm`),
- downloading ontology source files,
- building HPO, MONDO, ORPHANET, and SNOMED indexes,
- loading a FastHPOCR annotator from an index file, and
- converting annotation objects to R data frames with explicit offset columns.

## Installation

Install from GitHub with `remotes::install_github("sounkou-bioinfo/RfastHPOCR")`.

For ordinary use, `RfastHPOCR` follows current reticulate package practice:
`library(RfastHPOCR)` declares the Python requirements with
`reticulate::py_require()`, and reticulate can resolve them in an ephemeral
virtual environment when Python is first initialized.

```{r python-requirements}
library(RfastHPOCR)
fast_hpo_cr_python_packages()
fast_hpo_cr_available()
```

`FastHPOCR` 0.1.4 on PyPI does not currently declare all runtime dependencies,
so `RfastHPOCR` requests `pronto` and `tqdm` explicitly.

If you prefer a persistent named environment, create it and select it before
Python is initialized:

For a persistent environment, run `fast_hpo_cr_install(envname = "r-fast-hpo-cr", method = "virtualenv")`, restart R, and then call `fast_hpo_cr_use_env("r-fast-hpo-cr", method = "virtualenv")` before the first Python import.

To install from the upstream GitHub repository instead of PyPI:

To install the Python dependency from the upstream GitHub repository instead of PyPI, use `fast_hpo_cr_install(source = "github", ref = "main")`.

If Python was already initialized before selecting or installing an environment,
restart R so reticulate can bind to the intended environment.

## Ontology downloads

HPO is the primary target. HPO release assets are published at
<https://github.com/obophenotype/human-phenotype-ontology/releases>. The
`hp.obo` helper uses the stable OBO PURL and is the source file used by
FastHPOCR's HPO indexer. Additional HPO JSON, OWL, HPOA, and gene/phenotype
tables are exposed for validation and audit workflows.

```{r ontology-urls}
fast_hpo_cr_ontology_urls()[grep("^hpo_", names(fast_hpo_cr_ontology_urls()))]
```

For a real HPO analysis, use `download_hpo_obo("ontologies")` to fetch the
indexing source, plus optional audit resources such as `download_hpo_json()`,
`download_hpo_phenotype_hpoa()`, and `download_hpo_genes_to_phenotype()`.

MONDO and Orphanet download helpers are present for the other upstream
FastHPOCR indexers, but they are not the focus of this package. MONDO release
assets are published at <https://github.com/monarch-initiative/mondo/releases>.

The rest of this README is not a mock API sketch: it creates a tiny HPO-style
OBO file, runs the same HPO download helper through a local file source, indexes
it multiple ways with the Python FastHPOCR engine, annotates text, and writes
the annotation table. The local file source keeps README rendering fast and
network-independent while still exercising the real download, indexing,
re-indexing, annotation, and serialization code paths.

## End-to-end HPO indexing and annotation

Create a tiny HPO-like ontology fixture. It contains the HPO root needed by
FastHPOCR (`HP:0000118`), two top-level abnormality categories, and two concrete
phenotype terms.

```{r tiny-ontology}
work_dir <- file.path(tempdir(), "RfastHPOCR-readme")
unlink(work_dir, recursive = TRUE)
dir.create(work_dir, recursive = TRUE)

source_obo <- file.path(work_dir, "source-hp.obo")
writeLines(c(
  "format-version: 1.2",
  "ontology: hp",
  "",
  "[Term]",
  "id: HP:0000001",
  "name: All",
  "",
  "[Term]",
  "id: HP:0000118",
  "name: Phenotypic abnormality",
  "is_a: HP:0000001 ! All",
  "",
  "[Term]",
  "id: HP:0000119",
  "name: Abnormality of the genitourinary system",
  "is_a: HP:0000118 ! Phenotypic abnormality",
  "",
  "[Term]",
  "id: HP:0000707",
  "name: Abnormality of the nervous system",
  "is_a: HP:0000118 ! Phenotypic abnormality",
  "",
  "[Term]",
  "id: HP:0004322",
  "name: Short stature",
  "synonym: \"short height\" EXACT []",
  "is_a: HP:0000119 ! Abnormality of the genitourinary system",
  "",
  "[Term]",
  "id: HP:0001250",
  "name: Seizure",
  "synonym: \"seizures\" EXACT []",
  "is_a: HP:0000707 ! Abnormality of the nervous system"
), source_obo)

hp_obo <- download_hpo_obo(
  dest_dir = file.path(work_dir, "downloads"),
  url = source_obo,
  quiet = TRUE
)
basename(hp_obo)
```

Index both concrete phenotype branches and annotate a sentence.

```{r index-all}
index_all_log <- capture.output({
  index_all <- index_hpo(
    hpo_file = hp_obo,
    output_dir = file.path(work_dir, "index-all"),
    root_concepts = c("HP:0000119", "HP:0000707"),
    include_top_level_category = TRUE,
    compress_index = FALSE
  )
})
grep("Collected|Indexing HPO terms|Serializing index", index_all_log, value = TRUE)
file.exists(index_all)

ann_all <- hpo_annotator(index_all)
hits_all <- hpo_annotate(
  ann_all,
  "The patient has short stature and seizures.",
  longest_match = TRUE
)
hits_all[, c("span", "id", "label", "start", "end", "start_offset", "end_offset")]
```

Re-index the same ontology restricted to the nervous-system branch. This is a
real second FastHPOCR indexing pass; the genitourinary `Short stature` term is
not present in this restricted index.

```{r reindex-restricted}
index_neuro_log <- capture.output({
  index_neuro <- index_hpo(
    hpo_file = hp_obo,
    output_dir = file.path(work_dir, "index-neuro"),
    root_concepts = "HP:0000707",
    include_top_level_category = TRUE,
    compress_index = FALSE
  )
})
grep("Collected|Indexing HPO terms|Serializing index", index_neuro_log, value = TRUE)

ann_neuro <- hpo_annotator(index_neuro)
hpo_annotate(
  ann_neuro,
  "The patient has short stature and seizures.",
  longest_match = TRUE
)[, c("span", "id", "label")]
```

Re-index again with an external synonym file. The phrase `small stature` is not
in the tiny OBO file; it comes from the external synonym file passed through to
FastHPOCR.

```{r reindex-external-synonym}
syn_file <- file.path(work_dir, "external-synonyms.txt")
writeLines("HP:0004322=small stature", syn_file)

index_external_log <- capture.output({
  index_external <- index_hpo(
    hpo_file = hp_obo,
    output_dir = file.path(work_dir, "index-external"),
    root_concepts = c("HP:0000119", "HP:0000707"),
    external_syn_file = syn_file,
    include_top_level_category = TRUE,
    compress_index = FALSE
  )
})
grep("Collected|Indexing HPO terms|Serializing index", index_external_log, value = TRUE)

ann_external <- hpo_annotator(index_external)
hpo_annotate(ann_external, "Small stature was noted.")[, c("span", "id", "label")]
```

Serialize the R annotation table with the same tab-separated shape used by the
upstream Python package.

```{r write-annotations}
out_tsv <- file.path(work_dir, "annotations.tsv")
hpo_write_annotations(hits_all, out_tsv, include_categories = TRUE)
readLines(out_tsv)
```

## Role in the hybrid HPO extraction pipeline

The intended split is:

- **FastHPOCR**: deterministic extractor/linker baseline for explicit phenotype mentions.
- **Small local model**: phrase cleanup, paraphrase normalization, and candidate phenotype suggestions.
- **piknit / Pi**: auditable orchestration in R Markdown or Quarto, including local tool calls, model calls, and report generation.

In that design this package owns the deterministic FastHPOCR lane: text in,
candidate HPO rows out, with provenance and offsets preserved.

## Hybrid adjudication harness primitives

The package now also provides the small, model-agnostic pieces needed to compare
hybrid configurations without making `RfastHPOCR` depend on any one LLM provider.
The main practical arm is:

```text
note + FastHPOCR candidates -> model keep/drop adjudication
```

FastHPOCR still generates deterministic candidates. The model is asked only for
short auditable fields: `decision`, `evidence_span`, and `short_reason`.
Reasoning-token counts, when a provider exposes them, are stored only as run
metadata.

```{r harness-candidates}
review_text <- "Short stature was noted. No seizures were reported."
review_hits <- hpo_annotate(ann_all, review_text, longest_match = TRUE)
review_candidates <- hpo_candidate_table(review_hits, case_id = "readme-case-001")
review_candidates[, c("case_id", "candidate_id", "candidate_span", "hpo_id", "hpo_label")]
```

The harness modes are explicit so that `tool_only`, `model_only`, and hybrid
arms can be compared with the same term-level schema.

```{r harness-modes}
hpo_harness_modes()[, c("mode", "question")]
```

The prompt embeds the candidate rows and a strict JSON schema. The README prints
only the beginning of the prompt to keep the rendered document short.

```{r harness-prompt}
prompt <- hpo_adjudication_prompt(review_text, review_candidates)
cat(paste(head(strsplit(prompt, "\n", fixed = TRUE)[[1]], 16), collapse = "\n"))
```

A model response is expected to be JSON. This mocked response demonstrates the
clinical audit shape: keep direct patient findings and drop negated findings
with an evidence quote and a short reason.

```{r harness-parse}
mock_response <- jsonlite::toJSON(
  list(
    case_id = "readme-case-001",
    decisions = list(
      list(
        candidate_id = "readme-case-001:0001",
        candidate_span = "Short stature",
        normalized_phrase = "Short stature",
        hpo_id = "HP:0004322",
        hpo_label = "Short stature",
        decision = "keep",
        support_type = "direct",
        patient_context = "patient",
        evidence_span = "Short stature was noted",
        short_reason = "The phenotype is directly stated for the patient.",
        replacement_hpo_id = NULL,
        replacement_hpo_label = NULL,
        confidence = 0.95
      ),
      list(
        candidate_id = "readme-case-001:0002",
        candidate_span = "seizures",
        normalized_phrase = "seizures",
        hpo_id = "HP:0001250",
        hpo_label = "Seizure",
        decision = "drop",
        support_type = "none",
        patient_context = "negated",
        evidence_span = "No seizures were reported",
        short_reason = "The note explicitly negates seizures.",
        replacement_hpo_id = NULL,
        replacement_hpo_label = NULL,
        confidence = 0.99
      )
    )
  ),
  auto_unbox = TRUE,
  null = "null",
  pretty = TRUE
)

adjudicated <- hpo_parse_adjudication(mock_response, candidates = review_candidates)
adjudicated[, c(
  "candidate_span", "hpo_id", "decision", "patient_context",
  "evidence_span", "short_reason"
)]
```

`hpo_adjudicate_candidates()` accepts any model runner function. In a real
pipeline the runner can call `piknit::pi_run()`, Ollama, vLLM, LM Studio, or a
remote API. The run log captures provider/model metadata, token usage, latency,
parse success, and reasoning-token counts when available.

```{r harness-run-log}
mock_runner <- function(prompt) {
  out <- mock_response
  attr(out, "usage") <- list(
    input_tokens = 500L,
    output_tokens = 160L,
    total_tokens = 660L,
    reasoning_tokens = 0L,
    tool_call_count = 0L
  )
  out
}

run <- hpo_adjudicate_candidates(
  review_text,
  review_candidates,
  runner = mock_runner,
  provider = "mock-provider",
  model = "mock-small-model",
  run_id = "readme-run-001"
)
run$run_log[, c(
  "provider", "model", "mode", "input_tokens", "output_tokens",
  "reasoning_tokens", "latency_seconds", "parse_success"
)]
```

## Optional real-HPO clinical stress test

The tiny ontology above keeps the ordinary README fast and deterministic. For a
real HPO smoke test, set `RFASTHPOCR_README_REAL_HPO=true` or run
`make rdm-real-hpo`. That path downloads the real `hp.obo`, builds a real
FastHPOCR HPO index under `RFASTHPOCR_REAL_HPO_DIR` or a user cache directory,
and runs a context-heavy clinical note through the same harness primitives.

This is intentionally opt-in because a full HPO index is large and can take
several minutes to build, but the code below is real executable README code, not
pseudo-code.

```{r real-hpo-index, eval = real_hpo_readme}
real_hpo_dir <- Sys.getenv(
  "RFASTHPOCR_REAL_HPO_DIR",
  file.path(tools::R_user_dir("RfastHPOCR", "cache"), "real-hpo")
)
dir.create(real_hpo_dir, recursive = TRUE, showWarnings = FALSE)

real_hp_obo <- file.path(real_hpo_dir, "hp.obo")
if (!file.exists(real_hp_obo)) {
  real_hp_obo <- download_hpo_obo(real_hpo_dir, quiet = TRUE)
}

real_index_dir <- file.path(real_hpo_dir, "index")
real_index <- file.path(real_index_dir, "hp.index")
if (!file.exists(real_index)) {
  dir.create(real_index_dir, recursive = TRUE, showWarnings = FALSE)
  real_index_log <- capture.output({
    real_index <- index_hpo(
      real_hp_obo,
      real_index_dir,
      root_concepts = "HP:0000118",
      include_top_level_category = TRUE,
      compress_index = FALSE
    )
  })
} else {
  real_index <- normalizePath(real_index, mustWork = TRUE)
  real_index_log <- "Using cached real HPO index."
}

data.frame(
  hp_obo = basename(real_hp_obo),
  hpo_index = basename(real_index),
  index_size_mb = round(file.info(real_index)$size / 1024^2, 1),
  stringsAsFactors = FALSE
)
```

```{r real-hpo-stress-candidates, eval = real_hpo_readme}
stress_note <- paste(
  "The patient is a 6-year-old boy referred for genetics evaluation. He has global developmental delay, autism spectrum disorder, intellectual disability, brachydactyly, one hypomelanotic macule on the trunk, and one café-au-lait macule on the neck. Echocardiogram showed cardiomyopathy and a septal cardiac defect.",
  "He has no seizures, no hypotonia, no ataxia, and no hearing loss. There is no short stature and no microcephaly.",
  "Family history is notable for a brother with thrombocytosis and autism and a maternal aunt with cardiomyopathy. These findings are not present in the proband.",
  "Earlier notes described the child as floppy in infancy and not talking yet at 3 years of age. Teachers report that he learns slowly. Parents also describe occasional staring spells.",
  sep = "\n\n"
)

real_ann <- hpo_annotator(real_index)
stress_hits <- hpo_annotate(real_ann, stress_note, longest_match = TRUE)
stress_candidates <- hpo_candidate_table(
  stress_hits,
  case_id = "hpo-stress-001",
  source = "FastHPOCR-real-HPO"
)

stress_candidates[, c(
  "candidate_span", "hpo_id", "hpo_label", "start_offset", "end_offset"
)]
```

The real HPO candidate layer deliberately captures both true proband findings
and context traps: negated terms, family-history terms, and duplicate mentions.
That is exactly why the model layer should produce explicit keep/drop decisions
with an evidence quote.

```{r real-hpo-stress-adjudication, eval = real_hpo_readme}
negated_ids <- c(
  "HP:0001250", # seizures
  "HP:0001252", # hypotonia
  "HP:0001251", # ataxia
  "HP:0000365", # hearing loss
  "HP:0004322", # short stature
  "HP:0000252"  # microcephaly
)

negated_evidence <- c(
  "HP:0001250" = "no seizures",
  "HP:0001252" = "no hypotonia",
  "HP:0001251" = "no ataxia",
  "HP:0000365" = "no hearing loss",
  "HP:0004322" = "no short stature",
  "HP:0000252" = "no microcephaly"
)

stress_decision <- function(candidate) {
  hpo_id <- as.character(candidate$hpo_id)
  start <- as.integer(candidate$start_offset)

  if (hpo_id %in% negated_ids) {
    return(list(
      candidate_id = as.character(candidate$candidate_id),
      candidate_span = as.character(candidate$candidate_span),
      normalized_phrase = as.character(candidate$normalized_phrase),
      hpo_id = hpo_id,
      hpo_label = as.character(candidate$hpo_label),
      decision = "drop",
      support_type = "none",
      patient_context = "negated",
      evidence_span = unname(negated_evidence[[hpo_id]]),
      short_reason = "The note explicitly negates this phenotype for the proband.",
      replacement_hpo_id = NULL,
      replacement_hpo_label = NULL,
      confidence = 1
    ))
  }

  family_history <- hpo_id == "HP:0001894" ||
    (hpo_id == "HP:0000717" && start > 450L) ||
    (hpo_id == "HP:0001638" && start > 450L)

  if (family_history) {
    evidence <- if (hpo_id == "HP:0001894") {
      "brother with thrombocytosis"
    } else if (hpo_id == "HP:0000717") {
      "brother with thrombocytosis and autism"
    } else {
      "maternal aunt with cardiomyopathy"
    }
    return(list(
      candidate_id = as.character(candidate$candidate_id),
      candidate_span = as.character(candidate$candidate_span),
      normalized_phrase = as.character(candidate$normalized_phrase),
      hpo_id = hpo_id,
      hpo_label = as.character(candidate$hpo_label),
      decision = "drop",
      support_type = "none",
      patient_context = "family_history",
      evidence_span = evidence,
      short_reason = "The evidence is family history and is stated not to be present in the proband.",
      replacement_hpo_id = NULL,
      replacement_hpo_label = NULL,
      confidence = 1
    ))
  }

  list(
    candidate_id = as.character(candidate$candidate_id),
    candidate_span = as.character(candidate$candidate_span),
    normalized_phrase = as.character(candidate$normalized_phrase),
    hpo_id = hpo_id,
    hpo_label = as.character(candidate$hpo_label),
    decision = "keep",
    support_type = "direct",
    patient_context = "patient",
    evidence_span = as.character(candidate$candidate_span),
    short_reason = "The phenotype is directly stated for the proband.",
    replacement_hpo_id = NULL,
    replacement_hpo_label = NULL,
    confidence = 0.95
  )
}

stress_response <- jsonlite::toJSON(
  list(
    case_id = "hpo-stress-001",
    decisions = lapply(seq_len(nrow(stress_candidates)), function(i) {
      stress_decision(stress_candidates[i, , drop = FALSE])
    })
  ),
  auto_unbox = TRUE,
  null = "null"
)

stress_runner <- function(prompt) {
  out <- stress_response
  attr(out, "usage") <- list(
    input_tokens = NA_integer_,
    output_tokens = NA_integer_,
    total_tokens = NA_integer_,
    reasoning_tokens = NA_integer_,
    tool_call_count = 0L
  )
  out
}

stress_run <- hpo_adjudicate_candidates(
  stress_note,
  stress_candidates,
  runner = stress_runner,
  provider = "expected-output",
  model = "manual-stress-adjudicator",
  run_id = "hpo-stress-readme"
)

stress_run$adjudication[, c(
  "candidate_span", "hpo_id", "decision", "patient_context",
  "evidence_span", "short_reason"
)]
```

The optional real-HPO test also shows the boundary between the deterministic
candidate layer and the model-expansion arms: phrases such as "floppy in
infancy", "not talking yet", "learns slowly", and "staring spells" may require
`model_only` or `model_candidates_tools_model` runs if we want inferred or
paraphrased phenotype recovery.

## Development

This package follows the same lightweight R package workflow as the surrounding
packages in this workspace:

```sh
make rd             # roxygen2 docs + NAMESPACE
make test           # tinytest
make index-real-hpo # download/build a cached full HPO index
make rdm            # render README.md from README.Rmd
make rdm-real-hpo   # render README.md with the full real-HPO stress test
make site           # pkgdown site
make check          # R CMD check --as-cran --no-manual
```
