---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

embeddinggemma_weights <- Sys.getenv("EMBEDDING_GEMMA_WEIGHTS_FILE")
colbert_weights <- Sys.getenv("COLBERT_WEIGHTS_FILE")
bebel_weights <- Sys.getenv("BEBELM_WEIGHTS_FILE")
have_embeddinggemma <- file.exists(embeddinggemma_weights)
have_colbert <- file.exists(colbert_weights)
have_bebel <- file.exists(bebel_weights)
```

# ducksemantics

`ducksemantics` is a DuckDB-native semantic graph and evidence-grounding
package for R. It stores ontology nodes, aliases, graph edges, dense vectors,
and native ColBERT document token vectors in one auditable DuckDB database.

It uses the dedicated Rbebelm model surfaces rather than causal generator
hidden states:

- **EmbeddingGemma** provides dense, task-prompted vectors for broad retrieval
  and optional DuckDB HNSW candidate generation.
- **LFM2.5-ColBERT** provides exact late-interaction MaxSim reranking of stored
  candidate documents.
- **BebeLM** provides local structured judgment of deterministic candidates,
  including negation, uncertainty, family history, and subject context.

HPO, MONDO, ORPHANET, and local knowledge graphs are sources for the same graph
schema, not special-purpose APIs.

## Build a semantic graph

The bundled HPO fixture is a small, attribution-preserving subset of the real
Human Phenotype Ontology. It keeps the README executable without downloading an
ontology during rendering.

```{r graph}
library(ducksemantics)

conn <- ducksemantics_connect()
hpo <- ducksemantics_read_obo(
  system.file("extdata", "hpo-readme.obo", package = "ducksemantics", mustWork = TRUE),
  family = "HPO",
  source = "HPO README fixture"
)
ducksemantics_write_graph(
  conn,
  nodes = hpo$nodes,
  aliases = hpo$aliases,
  edges = hpo$edges,
  replace = TRUE
)

# Deterministic, explainable candidates. This intentionally includes negated
# mentions; BebeLM judgment decides whether the mention is a patient finding.
case_text <- "The proband has short stature and seizures but no diabetes mellitus."
mentions <- ducksemantics_annotate(
  conn,
  case_text,
  document_id = "case-001"
)
mentions
```

## Dense retrieval with EmbeddingGemma

Use a retrieval-trained dense encoder for broad candidate generation. Choose an
EmbeddingGemma task deliberately: vectors are only comparable when they use a
compatible prompt contract. This chunk executes a real native encoder when
`EMBEDDING_GEMMA_WEIGHTS_FILE` names a local GGUF.

```{r dense, eval=have_embeddinggemma}
library(Rbebelm)

embedding_model <- embeddinggemma_model_load(
  embeddinggemma_weights,
  num_threads = 2
)
dense_provider <- ducksemantics_embeddinggemma_provider(
  embedding_model,
  label = "embeddinggemma-hpo-v1",
  task = "semantic_similarity",
  dimensions = 256L
)

terms <- DBI::dbGetQuery(conn, "SELECT node_id, label FROM semantic_nodes WHERE family = 'HPO'")
vectors <- ducksemantics_embed_cached(
  terms$label,
  provider = dense_provider,
  cache_dir = file.path(tempdir(), "ducksemantics-embeddinggemma-hpo-cache")
)
ducksemantics_embedding_batch(
  vectors,
  subject_id = terms$node_id,
  subject_kind = "hpo_term",
  provider = "embeddinggemma-hpo-v1",
  text = terms$label
) |> ducksemantics_write_embeddings(conn, replace = TRUE)

# Exact DuckDB vector search is sufficient for this small example. An HNSW
# index is optional for a larger dense candidate collection.
dense_hits <- ducksemantics_embedding_query(
  ducksemantics_embed(dense_provider, "short height")[1L, ],
  provider = "embeddinggemma-hpo-v1",
  subject_kind = "hpo_term",
  top_k = 3L
) |> ducksemantics_embedding_search(conn)
dense_hits
```

## Exact late interaction with native ColBERT

ColBERT document vectors are persisted once. A query is encoded with its own
native query contract; the DuckDB reranker computes the exact MaxSim sum over
the stored document vectors. For a very large corpus, use dense HNSW, lexical,
or graph candidates first, then restrict `candidate_subject_id` for the MaxSim
pass. The chunk is evaluated during every render and runs the real native model
when `COLBERT_WEIGHTS_FILE` names a local GGUF.

```{r colbert}
if (have_colbert) {
  colbert_model <- colbert_model_load(colbert_weights, num_threads = 2)
  colbert_documents <- ducksemantics_colbert_provider(
    colbert_model,
    role = "document",
    label = "lfm2.5-colbert-350m-q4km"
  )

  ducksemantics_token_embedding_batch_from_provider(
    terms$label,
    provider = colbert_documents,
    subject_id = terms$node_id,
    subject_kind = "hpo_term"
  ) |> ducksemantics_write_token_embeddings(conn, replace = TRUE)

  hits <- ducksemantics_colbert_query(
    colbert_model,
    "developmental delay with seizures",
    provider = "lfm2.5-colbert-350m-q4km",
    subject_kind = "hpo_term",
    top_k = 25L
  ) |> ducksemantics_late_interaction_search(conn)
  hits
} else {
  message("Set COLBERT_WEIGHTS_FILE to execute the native ColBERT example.")
}
```

The token rows contain native 128-dimensional ColBERT document vectors, not
BebeLM causal states. `duckdb-vss` can accelerate the EmbeddingGemma dense
stage, but it does not implement variable-length ColBERT MaxSim.

## BebeLM candidate judgment

The final local LLM stage only receives candidates and context already grounded
by deterministic retrieval. Its structured response can drop negated,
uncertain, historical, or wrong-subject mentions; it may only replace a concept
with an explicitly supplied candidate or graph neighbor. This chunk executes a
real BebeLM call when `BEBELM_WEIGHTS_FILE` names a local GGUF.

```{r judge, eval=have_bebel}
generation_model <- bebel_model_load(bebel_weights, num_threads = 2)
agent <- bebel_agent(generation_model, greedy = TRUE, max_gen = 256L, max_think = 0L)

judgments <- ducksemantics_bebel_judge(
  agent,
  text = case_text,
  mentions = mentions,
  conn = conn,
  parser = ducksemantics_bebel_tool_judgment_parser(),
  model = "LFM2.5-8B-A1B"
)
judgments
```

```{r cleanup, include=FALSE}
DBI::dbDisconnect(conn, shutdown = TRUE)
```
