---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# RfastHPOCR

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/RfastHPOCR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/RfastHPOCR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`RfastHPOCR` is a reticulate-backed R binding to Tudor Groza's
[`FastHPOCR`](https://github.com/tudorgroza/fast_hpo_cr) Python package for
fast concept recognition over Human Phenotype Ontology terms in free text.

The package keeps the upstream Python implementation as the execution engine and
adds R-oriented helpers for:

- declaring and installing the Python dependency set (`FastHPOCR`, `pronto`, and
  `tqdm`),
- downloading ontology source files,
- building HPO, MONDO, ORPHANET, and SNOMED indexes,
- loading a FastHPOCR annotator from an index file, and
- converting annotation objects to R data frames with explicit offset columns.

## Installation

Install from GitHub with `remotes::install_github("sounkou-bioinfo/RfastHPOCR")`.

For ordinary use, `RfastHPOCR` follows current reticulate package practice:
`library(RfastHPOCR)` declares the Python requirements with
`reticulate::py_require()`, and reticulate can resolve them in an ephemeral
virtual environment when Python is first initialized.

```{r python-requirements}
library(RfastHPOCR)
fast_hpo_cr_python_packages()
fast_hpo_cr_available()
```

`FastHPOCR` 0.1.4 on PyPI does not currently declare all runtime dependencies,
so `RfastHPOCR` requests `pronto` and `tqdm` explicitly.

If you prefer a persistent named environment, create it and select it before
Python is initialized:

For a persistent environment, run `fast_hpo_cr_install(envname = "r-fast-hpo-cr", method = "virtualenv")`, restart R, and then call `fast_hpo_cr_use_env("r-fast-hpo-cr", method = "virtualenv")` before the first Python import.

To install from the upstream GitHub repository instead of PyPI:

To install the Python dependency from the upstream GitHub repository instead of PyPI, use `fast_hpo_cr_install(source = "github", ref = "main")`.

If Python was already initialized before selecting or installing an environment,
restart R so reticulate can bind to the intended environment.

## Ontology downloads

HPO is the primary target. HPO release assets are published at
<https://github.com/obophenotype/human-phenotype-ontology/releases>. The
`hp.obo` helper uses the stable OBO PURL and is the source file used by
FastHPOCR's HPO indexer. Additional HPO JSON, OWL, HPOA, and gene/phenotype
tables are exposed for validation and audit workflows.

```{r ontology-urls}
fast_hpo_cr_ontology_urls()[grep("^hpo_", names(fast_hpo_cr_ontology_urls()))]
```

For a real HPO analysis, use `download_hpo_obo("ontologies")` to fetch the
indexing source, plus optional audit resources such as `download_hpo_json()`,
`download_hpo_phenotype_hpoa()`, and `download_hpo_genes_to_phenotype()`.

MONDO and Orphanet download helpers are present for the other upstream
FastHPOCR indexers, but they are not the focus of this package. MONDO release
assets are published at <https://github.com/monarch-initiative/mondo/releases>.

The rest of this README is not a pseudo API sketch: it creates a tiny HPO-style
OBO file, runs the same HPO download helper through a local file source, indexes
it multiple ways with the Python FastHPOCR engine, annotates text, and writes
the annotation table. The local file source keeps README rendering fast and
network-independent while still exercising the real download, indexing,
re-indexing, annotation, and serialization code paths.

## End-to-end HPO indexing and annotation

Create a tiny HPO-like ontology fixture. It contains the HPO root needed by
FastHPOCR (`HP:0000118`), two top-level abnormality categories, and two concrete
phenotype terms.

```{r tiny-ontology}
work_dir <- file.path(tempdir(), "RfastHPOCR-readme")
unlink(work_dir, recursive = TRUE)
dir.create(work_dir, recursive = TRUE)

source_obo <- file.path(work_dir, "source-hp.obo")
writeLines(c(
  "format-version: 1.2",
  "ontology: hp",
  "",
  "[Term]",
  "id: HP:0000001",
  "name: All",
  "",
  "[Term]",
  "id: HP:0000118",
  "name: Phenotypic abnormality",
  "is_a: HP:0000001 ! All",
  "",
  "[Term]",
  "id: HP:0000119",
  "name: Abnormality of the genitourinary system",
  "is_a: HP:0000118 ! Phenotypic abnormality",
  "",
  "[Term]",
  "id: HP:0000707",
  "name: Abnormality of the nervous system",
  "is_a: HP:0000118 ! Phenotypic abnormality",
  "",
  "[Term]",
  "id: HP:0004322",
  "name: Short stature",
  "synonym: \"short height\" EXACT []",
  "is_a: HP:0000119 ! Abnormality of the genitourinary system",
  "",
  "[Term]",
  "id: HP:0001250",
  "name: Seizure",
  "synonym: \"seizures\" EXACT []",
  "is_a: HP:0000707 ! Abnormality of the nervous system"
), source_obo)

hp_obo <- download_hpo_obo(
  dest_dir = file.path(work_dir, "downloads"),
  url = source_obo,
  quiet = TRUE
)
basename(hp_obo)
```

Index both concrete phenotype branches and annotate a sentence.

```{r index-all}
index_all_log <- capture.output({
  index_all <- index_hpo(
    hpo_file = hp_obo,
    output_dir = file.path(work_dir, "index-all"),
    root_concepts = c("HP:0000119", "HP:0000707"),
    include_top_level_category = TRUE,
    compress_index = FALSE
  )
})
grep("Collected|Indexing HPO terms|Serializing index", index_all_log, value = TRUE)
file.exists(index_all)

ann_all <- hpo_annotator(index_all)
hits_all <- hpo_annotate(
  ann_all,
  "The patient has short stature and seizures.",
  longest_match = TRUE
)
hits_all[, c("span", "id", "label", "start", "end", "start_offset", "end_offset")]
```

Re-index the same ontology restricted to the nervous-system branch. This is a
real second FastHPOCR indexing pass; the genitourinary `Short stature` term is
not present in this restricted index.

```{r reindex-restricted}
index_neuro_log <- capture.output({
  index_neuro <- index_hpo(
    hpo_file = hp_obo,
    output_dir = file.path(work_dir, "index-neuro"),
    root_concepts = "HP:0000707",
    include_top_level_category = TRUE,
    compress_index = FALSE
  )
})
grep("Collected|Indexing HPO terms|Serializing index", index_neuro_log, value = TRUE)

ann_neuro <- hpo_annotator(index_neuro)
hpo_annotate(
  ann_neuro,
  "The patient has short stature and seizures.",
  longest_match = TRUE
)[, c("span", "id", "label")]
```

Re-index again with an external synonym file. The phrase `small stature` is not
in the tiny OBO file; it comes from the external synonym file passed through to
FastHPOCR.

```{r reindex-external-synonym}
syn_file <- file.path(work_dir, "external-synonyms.txt")
writeLines("HP:0004322=small stature", syn_file)

index_external_log <- capture.output({
  index_external <- index_hpo(
    hpo_file = hp_obo,
    output_dir = file.path(work_dir, "index-external"),
    root_concepts = c("HP:0000119", "HP:0000707"),
    external_syn_file = syn_file,
    include_top_level_category = TRUE,
    compress_index = FALSE
  )
})
grep("Collected|Indexing HPO terms|Serializing index", index_external_log, value = TRUE)

ann_external <- hpo_annotator(index_external)
hpo_annotate(ann_external, "Small stature was noted.")[, c("span", "id", "label")]
```

Serialize the R annotation table with the same tab-separated shape used by the
upstream Python package.

```{r write-annotations}
out_tsv <- file.path(work_dir, "annotations.tsv")
hpo_write_annotations(hits_all, out_tsv, include_categories = TRUE)
readLines(out_tsv)
```

## Role in the hybrid HPO extraction pipeline

The intended split is:

- **FastHPOCR**: deterministic extractor/linker baseline for explicit phenotype mentions.
- **Small local model**: phrase cleanup, paraphrase normalization, and candidate phenotype suggestions.
- **piknit / Pi**: auditable orchestration in R Markdown or Quarto, including local tool calls, model calls, and report generation. See <https://github.com/sounkou-bioinfo/piknit>.

In that design this package owns the deterministic FastHPOCR lane: text in,
candidate HPO rows out, with provenance and offsets preserved.

## Hybrid adjudication harness primitives

The package now also provides the small, model-agnostic pieces needed to compare
hybrid configurations without making `RfastHPOCR` depend on any one LLM provider.
The main practical arm is:

```text
note + FastHPOCR candidates -> model keep/drop adjudication
```

FastHPOCR still generates deterministic candidates. The model is asked only for
short auditable fields: `decision`, `evidence_span`, and `short_reason`.
Reasoning-token counts, when a provider exposes them, are stored only as run
metadata.

```{r harness-candidates}
review_text <- "Short stature was noted. No seizures were reported."
review_hits <- hpo_annotate(ann_all, review_text, longest_match = TRUE)
review_candidates <- hpo_candidate_table(review_hits, case_id = "readme-case-001")
review_candidates[, c("case_id", "candidate_id", "candidate_span", "hpo_id", "hpo_label")]
```

The harness modes are explicit so that `tool_only`, `model_only`, and hybrid
arms can be compared with the same term-level schema.

```{r harness-modes}
hpo_harness_modes()[, c("mode", "question")]
```

The prompt embeds the candidate rows and a strict JSON schema. The README prints
only the beginning of the prompt to keep the rendered document short.

```{r harness-prompt}
prompt <- hpo_adjudication_prompt(review_text, review_candidates)
cat(paste(head(strsplit(prompt, "\n", fixed = TRUE)[[1]], 16), collapse = "\n"))
```

A live model response is expected to be JSON. The real `piknit` / Pi call is
shown in the full-HPO stress test below; no hand-written adjudication response
is used in this README.

## Real-HPO clinical stress test

The tiny ontology above keeps the basic FastHPOCR example fast, but the real
harness should also be exercised against the current full HPO. The helper below
downloads `hp.obo` and builds a cached FastHPOCR index under
`RFASTHPOCR_REAL_HPO_DIR` or the R user cache directory. A fresh full-HPO build
can take several minutes; subsequent renders reuse the index.

```{r real-hpo-index}
real_index <- hpo_real_index()

data.frame(
  hp_obo = basename(attr(real_index, "hpo_file")),
  hpo_index = basename(real_index),
  index_size_mb = round(file.info(real_index)$size / 1024^2, 1),
  stringsAsFactors = FALSE
)
```

```{r real-hpo-stress-candidates}
stress_note <- paste(
  "The patient is a 6-year-old boy referred for genetics evaluation. He has global developmental delay, autism spectrum disorder, intellectual disability, brachydactyly, one hypomelanotic macule on the trunk, and one café-au-lait macule on the neck. Echocardiogram showed cardiomyopathy and a septal cardiac defect.",
  "He has no seizures, no hypotonia, no ataxia, and no hearing loss. There is no short stature and no microcephaly.",
  "Family history is notable for a brother with thrombocytosis and autism and a maternal aunt with cardiomyopathy. These findings are not present in the proband.",
  "Earlier notes described the child as floppy in infancy and not talking yet at 3 years of age. Teachers report that he learns slowly. Parents also describe occasional staring spells.",
  sep = "\n\n"
)

real_ann <- hpo_annotator(real_index)
stress_hits <- hpo_annotate(real_ann, stress_note, longest_match = TRUE)
stress_candidates <- hpo_candidate_table(
  stress_hits,
  case_id = "hpo-stress-001",
  source = "FastHPOCR-real-HPO"
)

stress_candidates[, c(
  "candidate_span", "hpo_id", "hpo_label", "start_offset", "end_offset"
)]
```

The real HPO candidate layer deliberately captures both true proband findings
and context traps: negated terms, family-history terms, and duplicate mentions.
That is exactly why the model layer should produce explicit keep/drop decisions
with an evidence quote.

```{r real-hpo-stress-adjudication}
pi_runner <- function(prompt) {
  usage <- list()
  reply <- piknit::pi_stream(
    prompt,
    provider = "openai-codex",
    model = "gpt-5.3-codex-spark",
    timeout = 300,
    on_delta = NULL,
    on_event = function(event) {
      if (identical(event$type, "turn_end") && !is.null(event$message$usage)) {
        usage <<- event$message$usage
      }
    }
  )
  attr(reply, "usage") <- usage
  reply
}

stress_run <- hpo_adjudicate_candidates(
  stress_note,
  stress_candidates,
  runner = pi_runner,
  provider = "openai-codex",
  model = "gpt-5.3-codex-spark",
  run_id = "hpo-stress-readme"
)

stress_run$adjudication[, c(
  "candidate_span", "hpo_id", "decision", "patient_context",
  "evidence_span", "short_reason"
)]

stress_run$run_log[, c(
  "provider", "model", "input_tokens", "output_tokens", "total_tokens",
  "reasoning_tokens", "estimated_cost_usd", "latency_seconds", "parse_success"
)]
```

The real-HPO test also shows the boundary between the deterministic
candidate layer and the model-expansion arms: phrases such as "floppy in
infancy", "not talking yet", "learns slowly", and "staring spells" may require
`model_only` or `model_candidates_tools_model` runs if we want inferred or
paraphrased phenotype recovery.

## Development

This package follows the same lightweight R package workflow as the surrounding
packages in this workspace:

```sh
make rd             # roxygen2 docs + NAMESPACE
make test           # tinytest
make index-real-hpo # download/build a cached full HPO index
make rdm            # render README.md with the full real-HPO + Pi stress test
make rdm-real-hpo   # alias for make rdm
make site           # pkgdown site
make check          # R CMD check --as-cran --no-manual
```
